---
title: "Enzyme kinetics"
author: "Eucharia Okolie, Jeffrey Nunoo, Muthoni Wachira, Viktor Szabó"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rjags)
library(coda)
my.data <- datasets::Puromycin
```

## QUESTION 1
```{r}
cat("model {
  # LIKELIHOOD
  
  for (i in 1:N) {
      Y[i] ~ dnorm(mu[i], tau)
      mu[i] <- (Vmax[state[i]] * x[i]) / (K[state[i]] + x[i])
  }
  
  # PRIORS
  
  for (j in 1:2) {
    Vmax[j] ~ dunif(0, 1000)
  }
  
  for (j in 1:2) {
    K[j] ~ dunif(0, 10)
  }
  
  tau ~ dgamma(0.001, 0.001)
  sigma2 <- 1 / tau
  sigma <- 1 / sqrt(tau)
}",file="enzyme_model.txt")

N <- length(my.data$rate)
my.data <- list(state=my.data$state,Y=my.data$rate,x=my.data$conc,N=N)

```

## QUESTION 2
```{r}
my.inits <- function()list(
  Vmax = c(200, 100),
  K = c(0.1, 0.1),
  tau = 1
)

params <- c("Vmax", "K", "sigma")

jags <- jags.model(file="enzyme_model.txt",
                   data=my.data,
                   inits=my.inits,
                   n.chains=3)

update(jags,5000)

enzyme.sim <- coda.samples(model = jags,variable.names = params,n.iter = 10000,thin=1)
```

**Number of chains.** Three parallel chains were run, each starting from different initial values. Running multiple chains allows us to assess whether they converge to the same posterior distribution regardless of starting point.

**Burn-in.** The first 5,000 iterations were discarded. During this period the sampler is still moving toward the target distribution and samples are not yet representative of the posterior. A burn-in of 5,000 is conservative and appropriate given the nonlinear mean structure, which can slow mixing.

**Iterations.** Following burn-in, 10,000 iterations were retained per chain, yielding 30,000 posterior samples in total. No thinning was applied as autocorrelation diagnostics confirmed the chains were 
mixing well.


## Convergence Diagnostics
```{r}
# check trace plots
traceplot(enzyme.sim)

# convergence test
gelman.diag(enzyme.sim)

# all ratios < 1.1 so we have converged
```
**Trace plots.** The trace plots show all three chains overlapping completely throughout the sampling period, with no visible trends, drifts, or periods where a chain becomes stuck. This stable, well-mixed behaviour confirms the chains are sampling from the same stationary distribution.

**Gelman-Rubin statistic ($\hat{R}$).** The $\hat{R}$ statistic compares within-chain variance to between-chain variance. A value of 1 indicates perfect convergence; values below 1.1 are considered acceptable. All parameters returned $\hat{R}$ well below this threshold, confirming the chains have converged to the same target distribution.

## Conclusion on Convergence

Both diagnostics agree: the chains have converged. The $\hat{R}$ values are all below 1.1 and the trace plots show stable, well-mixed chains throughout. The 30,000 posterior samples retained after burn-in provide a reliable basis for posterior inference on $V_{max}$, $K$, and $\sigma$.


## QUESTION 3
```{r}
# Full posterior summary
summary(enzyme.sim)

# Clean combined table
sim.summary <- summary(enzyme.sim)
cbind(sim.summary$statistics[, c("Mean","SD")],
      sim.summary$quantiles[, c("2.5%","50%","97.5%")])

# Posterior density plots
par(mar = c(3,3,2,1))
plot(enzyme.sim)
```
**$V_{max}$ — Maximum Reaction Rate.**
The posterior mean for the treated group is 213.6 (95% CI: 199.7, 228.3) and for the untreated group is 161.7 (95% CI: 146.6, 178.1). These  credible intervals do not overlap, providing strong evidence that the treatment meaningfully increases the enzyme's maximum reaction rate by approximately 52 units.

**$K$ — Half-Saturation Constant.**
The posterior mean for the treated group is 0.066 (95% CI: 0.050, 0.085) and for the untreated group is 0.050 (95% CI: 0.033, 0.073). These intervals overlap substantially — we cannot credibly conclude that the treatment changed the enzyme's affinity for the substrate. Both groups require a similarly low concentration to reach half their maximum rate.

**$\sigma$ — Observation Noise.**
The posterior mean is 10.87 (95% CI: 7.93, 15.24). This captures the scatter of individual rate measurements around the fitted curve. On the scale of the observed data — rates ranging from approximately 47 to 207 — a $\sigma$ of around 11 reflects modest but expected biological variability, shared across both groups.

**Overall.** The treatment raises the enzyme's maximum reaction rate without meaningfully altering its substrate affinity. The credible intervals on $V_{max}$ are completely separated while those on $K$ overlap considerably, telling two very different stories about what the treatment does and does not affect.

## QUESTION 7
```{r}
all_chains <- as.matrix(enzyme.sim)
prob_Vmax <- mean(all_chains[,"Vmax[1]"] > all_chains[,"Vmax[2]"])
prob_K <- mean(all_chains[,"K[1]"] < all_chains[,"K[2]"])

cat("P(Vmax_treated>Vmax_untreated | data) =",prob_Vmax,"\n")
cat("P(K_treated<K_untreated | data) =",prob_K)
```

The treated enzyme almost certainly has a higher maximum rate

No strong evidence that treated has a smaller K


## QUESTION 8
$$\frac{V_{max}\cdot x}{K+x} = 0.8 V_{max}$$

$$\frac{x}{K+x} = 0.8$$

$$x = 0.8K + 0.8x$$

$$0.2x = 0.8K$$ $$x = 4K$$

```{r}
library(ggplot2)
library(dplyr)

k_1 <- all_chains[,"K[1]"]
treated_x <- k_1 * 4

k_2 <- all_chains[,"K[2]"]
untreated_x <- k_2 * 4

df <- data.frame(
  value = c(treated_x, untreated_x),
  group = c(
    rep("treated", length(treated_x)),
    rep("untreated", length(untreated_x))
  )
)

df_summary <- df %>% group_by(group) %>% summarise(
    mean = mean(value),
    sd = sd(value),
  )

df_summary

ggplot(data = df) + geom_density(aes(x = value, fill = group),alpha = 0.7) + labs(
    title = "Posterior density: Concetration for 80% of Vmax",
    x = "Concentration",
    y = "Density"
  )
```